Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-07-25 ms.tex
Some changes to DM paragraph in section 1. Other minor tweaks

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields in the cluster. The relative cosmic-ray pressure $X_{\mathrm{CR}}$ is. The observations imply a lower limit on the intra-cluster magnetic field of X$\mu$G. Finally, a constraint on the dark matter annihilation cross section is derived, $\expval{\sigma v}=$.
\end{abstract}

@@

\section{Introduction}
Clusters of galaxies are the largest observable structures in the Universe, with typical diameters of several Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. These gravitationally bound objects have formed on increasingly larger scales during the evolution of the Universe through the process of cosmic structure formation \citep{article:Voit:2005}. Although most of the cluster mass comes from large amounts of dark matter embedded in a cluster, as indicated by cluster dynamics and weak gravitational lensing \citep[][]{article:DiaferioSchindlerDolag:2008}, gas making up the intra-cluster medium (ICM) contributes about 15\% of the total cluster mass. Individual galaxies contribute only a small amount (about 5\%) to the cluster mass. The ICM gas mass also comprise a significant fraction of the observable (baryonic) matter in the Universe.

@@

\paragraph{rewrite this}
Depending on the CR energy density, quantified by the fraction of energy in CRs to the thermal energy in a cluster, $\eta=\expval{ E_{\mathrm{CR}}/E_{\mathrm{th}}}$, an observable flux of high-energy gamma rays can be expected. This fraction could plausibly be as large as 50\% \citep{article:Ryu_etal:2003}, but a much smaller fraction of only a few percent is required in order to produce a gamma-ray flux observable with current generation gamma-ray telescopes.

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a dark matter particle, e.g. a weakly interacting massive particle (WIMP). 
%This is particularly compelling because of the large amount of dark matter in a galaxy cluster.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations. In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_{0} = 100h$ km s$^{-1}$ Mpc$^{-1}$ with $h=0.7$ has been used.

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The move had no impact on the Coma cluster data set.}

@@

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core.}
\tablehead{
\colhead{Spatial model} & \multicolumn{3}{c}{S[$\sigma$]} & \multicolumn{3}{c}{Flux U.L.} \\
\colhead{} & \colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} & 
\colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} \\
\hline
\multicolumn{7}{c}{Spectral index $\alpha=2.1$ (fixed)}
}
\startdata
Point source & 0.30 & 0 & 1.3 & 1.73 & 4.87 & 0.627 \\
Disk, $r=0.2^{\circ}$ & 0.65 & 0.22 & 2.0 & 2.03 & 0.691 & 0.783 \\
Disk, $r=0.4^{\circ}$ & 0.77 & 0.18 & 2.3 & 2.39 & 0.757 & 0.847 \\
\cutinhead{Spectral index $\alpha=2.5$ (fixed)}
Point source & 0.33 & 0 & 1.42 & 1.81 & 0.526 & 0.635 \\
Disk, $r=0.2^{\circ}$ & 0.66 & 0.35 & 2.1 & 2.12 & 0.748 & 0.785 \\
Disk, $r=0.4^{\circ}$ & 0.80 & 0.36 & 2.4 & 2.50 & 0.833 & 0.849 \\
\enddata
\label{table:fermi}
\end{deluxetable}

>>>>

The gamma-ray flux upper limits have been used to constrain the populations of cosmic-ray particles and magnetic fields in the cluster. The relative cosmic-ray pressure $X_{\mathrm{CR}}$ is less than\ldots. The observations imply a lower limit on the intra-cluster magnetic field on the order of a $\mu$G depending on the magnetic decline and gamma-ray spectral index. Finally, a constraint on the thermally averaged product of the total self-annihilation cross section and the relative velocity of the dark-matter particles is derived, $\expval{\sigma v}\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}>X$ GeV/c$^{2}$.
\end{abstract}

@@

\section{Introduction}
Clusters of galaxies are the largest observable structures in the Universe, with typical diameters of several Mpc and masses on the order of $10^{14}$ to $10^{15} M_{\odot}$. These gravitationally bound objects have formed on increasingly larger scales during the evolution of the Universe through the process of cosmic structure formation \citep{article:Voit:2005}. Although most of the cluster mass comes from large amounts of dark matter (DM) embedded in a cluster, as indicated by cluster dynamics and weak gravitational lensing \citep[][]{article:DiaferioSchindlerDolag:2008}, gas making up the intra-cluster medium (ICM) contributes about 15\% of the total cluster mass. Individual galaxies contribute only a small amount (about 5\%) to the cluster mass. The ICM gas mass also comprise a significant fraction of the observable (baryonic) matter in the Universe.

@@

\paragraph{Rewrite this paragraph!}
Depending on the CR energy density, quantified by the fraction of energy in CRs to the thermal energy in a cluster, $\eta=\expval{ E_{\mathrm{CR}}/E_{\mathrm{th}}}$, an observable flux of high-energy gamma rays can be expected. This fraction could plausibly be as large as 50\% \citep{article:Ryu_etal:2003}, but a much smaller fraction of only a few percent is required in order to produce a gamma-ray flux observable with current generation gamma-ray telescopes.

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the DM density along the line of sight, making clusters of galaxies potentially good candidates for DM searches.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thouroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest and also one of the most massive clusters \citep[$M \sim 10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and succesively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo. As such it is a natural candidate for gamma-ray observations.

In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, 
together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_{0} = 100h$ km s$^{-1}$ Mpc$^{-1}$ with $h=0.7$ has been used.

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The move had no impact on the this data set.}

@@

\begin{deluxetable}{lcccccc}
\tablecolumns{7}
\tablewidth{0pc}
\tablecaption{\emph{Fermi}-LAT upper limits for the Coma cluster core.}
\tablehead{
\colhead{Spatial model} & \multicolumn{3}{c}{S[$\sigma$]\tablenotemark{a}} & \multicolumn{3}{c}{Flux U.L.\tablenotemark{b}} \\
\colhead{} & \colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} & 
\colhead{1-3 GeV} & \colhead{3-10 GeV} & \colhead{10-30 GeV} \\
\hline
\multicolumn{7}{c}{Spectral index $\alpha=2.1$ (fixed)}
}
\startdata
Point source & 0.30 & 0 & 1.3 & 1.73 & 4.87 & 0.627 \\
Disk, $r=0.2^{\circ}$ & 0.65 & 0.22 & 2.0 & 2.03 & 0.691 & 0.783 \\
Disk, $r=0.4^{\circ}$ & 0.77 & 0.18 & 2.3 & 2.39 & 0.757 & 0.847 \\
\cutinhead{Spectral index $\alpha=2.5$ (fixed)}
Point source & 0.33 & 0 & 1.42 & 1.81 & 0.526 & 0.635 \\
Disk, $r=0.2^{\circ}$ & 0.66 & 0.35 & 2.1 & 2.12 & 0.748 & 0.785 \\
Disk, $r=0.4^{\circ}$ & 0.80 & 0.36 & 2.4 & 2.50 & 0.833 & 0.849 \\
\enddata
\tablenotetext{a}{Statistical significance.}
\tablenotetext{b}{95\% confidence level upper limit in units of $10^{-10}$ ph. cm$^{-2}$ s$^{-1}$.}
\label{table:fermi}
\end{deluxetable}

%\begin{deluxetable}{lccc}
%\tablecolumns{4}
%\tablewidth{0pc}
%\tablecaption{Limits on the DM annihilation cross section $\expval{\sigma v}$.}
%\tablehead{
%\colhead{Channel} & RoI [deg] & $m_{\chi}$ [GeV] & $\expval{\sigma v}$ [cm$^{3}$ s$^{-1}$] \\
%}
%\startdata
%W$^{+}$W$^{-}$ & & & \\
%$b\bar{b}$ & & & \\
%$\tau^{+}\tau^{-}$ & & & \\
%\enddata
%\tablecomments{Limits are for different integration regions and DM particle mass $m_{\chi}$ derived from %the VERITAS gamma-ray flux upper limits presented in this work.}
%\end{deluxetable}
